Reproducible Research - Peer Assessment 1
=========================================

### Loading and Preprocessing ###
Assuming that the Activity Monitoring Data has been downloaded and unzipped to the current working directory, the file is read into the **activity** object, with the format specified for each column.

```{r, loadandprocess, echo=TRUE}
activity <- read.csv("activity.csv", colClasses = c("numeric", "Date", "numeric"));
```

### Mean Total Number of Steps Taken Per Day ###
For this part of the assignment, I first aggregate steps taken by date (summing over all intervals in each day), and display the results in a histogram showing the total number of steps taken each day.  
```{r, meantotal, echo=TRUE}
aggdata <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)
hist(aggdata$x, main="Total Number of Steps Per Day", xlab="Steps")
```
The **mean** total number of steps taken per day = *mean(aggdata$x)* = **`r format(mean(aggdata$x), digits=4) `**

The **median** total number of steps taken per day = *median(aggdata$x)* = **`r format(median(aggdata$x), digits=5) `**

### Average Daily Activity Pattern ###
The average daily pattern can be established by taking the aggregate of the mean number of steps taken in each interval. The results are displayed in a time series plot.
```{r, dailyaverage, echo=TRUE}
aggdata2 <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
plot(aggdata2$Group.1, aggdata2$x, xlab= "Intervals", ylab= "Steps", type='l', col='red')
```

This allows me to identify the 5-minute interval which contains, on average, the maximum number of steps.
```{r, max, echo=TRUE}
maxrow = subset(aggdata2[aggdata2$x == max(aggdata2$x),])
maxinterval <- as.numeric(maxrow[1])
```
Interval with the maximum number of steps = **`r maxinterval `**

### Imputing Missing Values ###
The number of rows containing missing values can be calculated using the **complete.cases** function.

```{r, missingvalues, echo=TRUE}
ok <- complete.cases(activity)
```
Number of rows with missing values = *sum(!ok)* = **`r sum(!ok) `**

My strategy for filling in these values was to use the mean for a particular interval. I chose not to differentiate between weekends and weekdays, for the sake of simplicity, and because there was no reason, at this point, to suspect a difference between the two. (Also, without knowing where this data was collected, one cannot assume that the weekend corresponds to Saturday and Sunday!) 
```{r, missingvalues1, echo=TRUE}
for (i in which(sapply(activity, is.numeric))) {
  activity[is.na(activity[, i]), i] <- mean(activity[, i],  na.rm = TRUE)
}
aggdata3 <- aggregate(activity$steps, by=list(activity$date), FUN=sum)
```

The histogram below shows that the distribution is slightly different with the missing values filled in. This is held up by the new **mean** and **median** values also, which are, as expected, slightly higher now that missing values have been filled in. 
```{r, missingvalues2, echo=TRUE}
hist(aggdata3$x, main="Total Number of Steps Per Day", xlab="Steps")
```
The **mean** total number of steps taken per day = mean(aggdata3$x) = **`r format(mean(aggdata3$x), digits=4) `**

The **median** total number of steps taken per day = median(aggdata3$x) = **`r format(median(aggdata3$x), digits=5) `**

### Differences Between Weekdays and Weekends ###
As suggested, I used the **weekdays()** function to create a factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or a weekend day.
```{r, weekdays, echo=TRUE}
activity$weektime <- as.factor(ifelse(weekdays(activity$date) %in% c("Saturday","Sunday"),"weekend", "weekday"))
```

I then aggregate the mean number of steps by interval and "weektime" (i.e. whether the date is a weekday or weekend day). This gives me 576 results (288 weekday intervals plus 288 weekend intervals), which I can then display in panel plots using the lattice system. One can see quite clearly a difference between weekday and weekend activity.
```{r, weekdays1, echo=TRUE}
aggdata4 <- aggregate(activity$steps, by=list(activity$interval, activity$weektime), FUN=mean)
library(lattice);
xyplot(x ~ Group.1 | Group.2, data=aggdata4, layout=c(1,2), type="l");
```

That's all for now, folks! :)